\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-10 jasa/paper-jasa.tex

The first fact to highlight is that, in the finite sample setting, uncalibrated
direct approximations $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat
s(\mathbf{x})} \right)$, as shown in green, may not always yield very accurate
estimates of the ratio, even when using universal approximators such as a neural
network or a random forest. This observation is at the origin of this work, and
led us to show that the ratio $r(\mathbf{x})$ can still be computed exactly by
forming the equivalent statistic
$\frac{p(s(\mathbf{x})|\gamma=0.05)}{p(s(\mathbf{x})|\gamma=0)}$, provided the
transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
constructing $\frac{\hat p(\hat s(\mathbf{x})|\gamma=0.05)}{\hat p(\hat
s(\mathbf{x})|\gamma=0)}$ as we propose, estimates appear much closer to the
true ratio than their direct counterparts, as illustrated in red. The
approximated ratios show to be more accurate for all three classifiers, even
when the direct approximation was not very good in the first place, e.g. as for the
linear model. In regions where the densities are low (i.e., at the
boundaries) the approximated ratio show to be less acurate since the underlying
classifier cannot make precise predictions, because of the few training samples
it is trained on in these parts of the input space. When leveraging the fact
that densities are mixtures and decomposing their ratio, the capacity of the
underlying classifiers can be focused on easier supervised learning tasks,
resulting as expected in even more accurate approximations, as shown in cyan.

As shown in the figures, qualitatively different results are obtained depending
the classification algorithm embedded within the approximation of the ratio. The
neural network-based approximation appears to work best, due to the capacity of
the classifier to model a smooth decision function $s(\mathbf{x})$ which is
very close to the optimal decision function. The random forest yields a decision function
which is piece-wise constant, therefore non-continuous and not strictly
monotonic with the ratio in the finite sample regime. On this problem, the
random forest still approximates the ratio correctly, but more variance can be
observed. Finally, the linear model produces a decision function that is
smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
the ratio around $\mathbf{x}=1$, this still results in an accurate
approximation, but poorer results should be expected on more complex problems.
While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
s(\mathbf{x}))$ is key to obtain accurate results. Standard
histograms with fixed bins have been used here for illustrative purposes, but we
anticipate that the embedded density estimation algorithm should be carefully
chosen and tuned, on a case by case basis. Overall, this first example confirms
that even if the classifier does a poor joob at modeling the optimal decision
function, good approximations of the likelihood ratio can still be obtained
with the proposed method, by calibrating $s(\mathbf{x})$ and/or by decomposing
the mixture when possible.

@@

    \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator $\hat \gamma$, as empirically estimated 1000 times over resampled observed data.}
    \label{fig:2}
\end{figure}

Given observed data from Figure~\ref{fig:1a}, let us now consider the composite hypothesis testing setting, by comparing
$H_0: \gamma = 0.05$ versus the alternative $H_1: \gamma \neq 0.05$. That is, we want to evaluate
\begin{equation}
\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
\end{equation}
By exploiting the fact that
\begin{equation}
\log \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} = \log \frac{ p({\cal D} | \gamma = 0.05) }{ p({\cal D} | \gamma = 0 ) } - \log \frac{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma) }{  p({\cal D} | \gamma = 0 ) },
\end{equation}
the generalized likelihood ratio statistic can be computed by evaluating
both terms reusing the decomposed neural network-based approximation from Figure~\ref{fig:1c}.
In particular, the maximum likelihood estimator (MLE) in the second term
can be found as outlined in
Section~\ref{sec:generalized-likelihood-ratio}.
Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
compared to the distribution of the true MLEs. It clearly suggests that
the proposed inference procedure yields in this case an unbiased estimator
($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
trials). Note that had the underlying ratio not been accurate enough because of
poor training or poor calibration, the two distributions would not be exactly
overlapping, indicating bias. Figure \ref{fig:2b} also confirms
that $-2 \log \Lambda$, as approximated with our method, is asymptotically $\chi^2_1$ distributed,
as should be the true statistic when the null is true.

@@

Let us now consider the more challenging problem of maximum likelihood
inference from multidimensional data generated by a black-box simulator.
For the sake of the illustration, we will assume 5-dimensional observed samples $\mathbf{x}$ generated
from the following process $p_0$:
    \begin{enumerate}
        \item $\mathbf{z} := (z_0, z_1, z_2, z_3, z_4)$, such that
            $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
            $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
            $z_2 \sim {\text{Mixture}}({\cal N}(\mu=-2, \sigma=1), {\cal N}(\mu=2, \sigma=0.5))$,
            $z_3 \sim {\text{Exponential}(\lambda=3)}$ and
            $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
        \item $\mathbf{x} := R  \mathbf{z}$, where $R$ is a fixed
        semi-positive definite $5 \times 5$ matrix defining a fixed projection
        of $\mathbf{z}$ into the observed space.
    \end{enumerate}
Given observations ${\cal D}$ represented in Figure~\ref{fig:4},
our goal is to infer the nominal values $\alpha=1$ and $\beta=-1$ that
were used for generating this data.

\begin{figure}
    \centering
    \includegraphics[width=1.0\textwidth]{figures/fig4.pdf}
    \caption{Observed 5-dimensional data ${\cal D}$ for nominal values $(\alpha=1, \beta=-1)$.}
    \label{fig:4}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[height=12.5em]{figures/fig5a.pdf}
        \caption{True mean negative log-ratio}
        \label{fig:5a}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[height=12.5em]{figures/fig5b.pdf}
        \caption{Estimated mean negative log-ratio}
        \label{fig:5b}
    \end{subfigure}
    \caption{Inference from parameterized approximated likelihood ratios.
             (\ref{fig:5a}) True mean negative log-ratio for the observed data ${\cal D}$. The red dot corresponds
             to the nominal values $(\alpha=1, \beta=-1)$, while the green dot is the maximum likelihood estimator
             $(\hat \alpha=1.006, \hat \beta=-1.012)$.
             (\ref{fig:5b}) Estimated mean negative log-ratio, as approximated from the constructed objective by
             the Gaussian process within the Bayesian optimization procedure. The blue dot is the estimated maximum
             likelihood estimator $(\hat \alpha=0.956, \hat \beta=-0.869)$ while the white dots correspond to
             the points sampled during optimization.}
    \label{fig:5}
\end{figure}

As previously, we make use of Eqn.~\ref{eq:mle_withs} for parameter inference
with approximated likelihood ratios and fix the alternate to
$\theta_1:=(\alpha=0, \beta=0)$. Following recommendations from
Section~\ref{sec:param-clf} to reduce computational requirements, we train a
single parameterized classifier $s(\mathbf{x};\theta:=(\alpha,\beta),\theta_1)$;
in this case, a 2-layer neural network. Since the generative model is not
expensive, the classifier output is calibrated on-the-fly with isotonic
regression for every candidate parameter pair $\theta:=(\alpha,\beta)$.
Minimizing the mean negative log of the constructed ratio with Bayesian
optimization~\citep{brochu2010tutorial} eventually leads to the estimated
posterior mean shown in Figure~\ref{fig:5b}, which is close to the true
objective illustrated in Figure~\ref{fig:5a}. After $50$ iterations, the
estimated minimizer is found at $(\hat \alpha=0.956, \hat \beta=-0.869$). Again,
repeated experiments (omitted here) confirm that the proposed procedure yields
an unbiased maximum likelihood estimator.

>>>>

The first fact to highlight is that, in the finite sample setting, uncalibrated
direct approximations $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat
s(\mathbf{x})} \right)$, as shown in green, may not always yield very accurate
estimates of the ratio, even when using universal approximators such as a neural
network or a random forest. This observation is at the origin of this work, and
led us to show that the ratio $r(\mathbf{x})$ can still be computed exactly by
forming the equivalent statistic
$\frac{p(s(\mathbf{x})|\gamma=0.05)}{p(s(\mathbf{x})|\gamma=0)}$, provided the
transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
constructing $\frac{\hat p(\hat s(\mathbf{x})|\gamma=0.05)}{\hat p(\hat
s(\mathbf{x})|\gamma=0)}$ as we propose, estimates appear much closer to the
true ratio than their direct counterparts, as illustrated in red. The
approximated ratios show to be more accurate for all three classifiers, even
when the direct approximation was not very good in the first place, e.g. as for the
linear model. In regions where the densities are low (i.e., at the
boundaries) the approximated ratio show to be less accurate since the underlying
classifier cannot make precise predictions, because of the few training samples
it is trained on in these parts of the input space. When leveraging the fact
that densities are mixtures and decomposing their ratio, the capacity of the
underlying classifiers can be focused on easier supervised learning tasks,
resulting as expected in even more accurate approximations, as shown in cyan.

As shown in the figures, qualitatively different results are obtained depending
the classification algorithm embedded within the approximation of the ratio. The
neural network-based approximation appears to work best, due to the capacity of
the classifier to model a smooth decision function $s(\mathbf{x})$ which is
very close to the optimal decision function. The random forest yields a decision function
which is piece-wise constant, therefore non-continuous and not strictly
monotonic with the ratio in the finite sample regime. On this problem, the
random forest still approximates the ratio correctly, but more variance can be
observed. Finally, the linear model produces a decision function that is
smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
the ratio around $\mathbf{x}=1$, this still results in an accurate
approximation, but poorer results should be expected on more complex problems.
While not directly apparent, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
s(\mathbf{x}))$ is key to obtain accurate results. Standard
histograms with fixed bins have been used here for illustrative purposes, but we
anticipate that the embedded density estimation algorithm should be carefully
chosen and tuned, on a case by case basis. Overall, this first example confirms
that even if the classifier does a poor job at modeling the optimal decision
function, good approximations of the likelihood ratio can still be obtained
with the proposed method, by calibrating $s(\mathbf{x})$ and/or by decomposing
the mixture when possible.

@@

    \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator
            $\hat \gamma$, as empirically estimated 1000 times over re-sampled observed data.}
    \label{fig:2}
\end{figure}

Given observed data from Figure~\ref{fig:1a}, let us now consider the composite hypothesis testing setting, by comparing
$H_0: \gamma = 0.05$ versus the alternative $H_1: \gamma \neq 0.05$. That is, we want to evaluate
\begin{equation}
\Lambda({\cal D}; \Theta_0 = \{0.05\}, \Theta=[0, 1]) =  \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} \; .
\end{equation}
By exploiting the fact that
\begin{equation}
\log \frac{ p({\cal D} | \gamma = 0.05) }{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma )} = \log \frac{ p({\cal D} | \gamma = 0.05) }{ p({\cal D} | \gamma = 0 ) } - \log \frac{ \sup_{\gamma \in \Theta} p({\cal D} | \gamma) }{  p({\cal D} | \gamma = 0 ) },
\end{equation}
the generalized likelihood ratio statistic can be computed by evaluating
both terms reusing the decomposed neural network-based approximation from Figure~\ref{fig:1c}.
In particular, the maximum likelihood estimator (MLE) in the second term
can be found as outlined in
Section~\ref{sec:generalized-likelihood-ratio}.
Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
compared to the distribution of the true MLEs. It clearly suggests that
the proposed inference procedure yields in this case an unbiased estimator
($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
trials). Note that had the underlying ratio not been accurate enough because of
poor training or poor calibration, the two distributions would not be exactly
overlapping, indicating bias. From this, the initial $\Lambda$ test statistic can finally
be evaluated to determine in this case that the null should not be rejected.  Figure \ref{fig:2b} also confirms
that $-2 \log \Lambda$, as approximated with our method, is asymptotically $\chi^2_1$ distributed,
as should be the true statistic when the null is true.

@@

Let us now consider the more challenging problem of maximum likelihood
inference from multidimensional data generated by a black-box simulator.
For the sake of the illustration, we will assume 5-dimensional observed samples $\mathbf{x}$ generated
from the following process $p_0$:
    \begin{enumerate}
        \item $\mathbf{z} := (z_0, z_1, z_2, z_3, z_4)$, such that
            $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
            $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
            $z_2 \sim {\text{Mixture}}({\cal N}(\mu=-2, \sigma=1), {\cal N}(\mu=2, \sigma=0.5))$,
            $z_3 \sim {\text{Exponential}(\lambda=3)}$ and
            $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
        \item $\mathbf{x} := R  \mathbf{z}$, where $R$ is a fixed
        semi-positive definite $5 \times 5$ matrix defining a fixed projection
        of $\mathbf{z}$ into the observed space.
    \end{enumerate}

\begin{figure}
    \centering
    \includegraphics[width=1.0\textwidth]{figures/fig4.pdf}
    \caption{Observed 5-dimensional data ${\cal D}$ for nominal values $(\alpha=1, \beta=-1)$. The size of ${\cal D}$ is set to $500$.}
    \label{fig:4}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}[t]{0.31\textwidth}
        \includegraphics[width=\textwidth]{figures/fig5a.pdf}
        \caption{  }
        \label{fig:5a}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.31\textwidth}
        \includegraphics[width=\textwidth]{figures/fig5b.pdf}
        \caption{ }
        \label{fig:5b}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.31\textwidth}
        \includegraphics[width=\textwidth]{figures/fig5c.pdf}
        \caption{ }
        \label{fig:5c}
    \end{subfigure}
    \caption{Inference from parameterized approximated likelihood ratios. The red dot corresponds
    to the nominal values $(\alpha=1, \beta=-1)$, the green dot is the maximum likelihood estimator
    $(\hat \alpha=0.896, \hat \beta=-0.984)$ while the blue dot is the estimated maximum likelihood estimator $(\hat \alpha=0.893, \hat \beta=-1.135)$.
    1, 2 and 3-$\sigma$ contours are shown in white.
             (\ref{fig:5a}) Exact values of the objective for the observed data ${\cal D}$.
             (\ref{fig:5b}) Sampled values of the approximated objective for a $15 \times 15$ fixed grid.
             (\ref{fig:5c}) Posterior mean of the approximated objective, as estimated by the Gaussian process within the Bayesian optimization procedure. White dots show the sampled candidates during the optimization process.}
    \label{fig:5}
\end{figure}

Given observations ${\cal D}$ represented in Figure~\ref{fig:4},
our goal is to infer the nominal values $\alpha=1$ and $\beta=-1$ that
were used for generating this data.
We use the objective function
\begin{equation}\label{eqn:obj-ex2}
    -2 \log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }
\end{equation}
that we approximate by exploiting the fact that
\begin{equation}\label{eqn:obj-ex2-expanded}
\log \frac{p({\cal D} | \alpha, \beta)}{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) } = \log \frac{p({\cal D} | \alpha, \beta)}{ p({\cal D} | \alpha=0, \beta=0) } - \log \frac{ \sup_{\alpha,\beta} p({\cal D} | \alpha, \beta) }{ p({\cal D} | \alpha=0, \beta=0)}.
\end{equation}
Following recommendations from Section~\ref{sec:param-clf} to reduce
computational requirements, both terms in the right hand side of Eqn.~\ref{eqn:obj-ex2-expanded} are evaluated using a single
parameterized approximated  likelihood ratio, in this case based on a parameterized
2-layer neural network where the alternate is fixed to $\alpha=0, \beta=0$.
Since the generative model is not expensive, the classifier output is calibrated
on-the-fly with histograms for  every candidate parameter pair $(\alpha,\beta)$.
Minimizing the constructed approximation of the objective with
Bayesian optimization~\citep{brochu2010tutorial} eventually leads to the
estimated posterior mean shown in Figure~\ref{fig:5c}, which is close to the
true objective illustrated in Figure~\ref{fig:5a}. After $50$ iterations, the
estimated minimizer is found at $(\hat \alpha=0.893, \hat \beta=-1.135$). For
comparison, Figure~\ref{fig:5b} shows the objective value as directly
approximated by the proposed method for a fixed grid of values. Since $\hat s(\mathbf{x;\alpha,\beta})$ is calibrated on-the-fly,
roughness in the contours can be observed, in particular due to non-smooth changes in the resulting histograms as $\alpha$ and $\beta$ vary.
As discussed in Section~\ref{sec:param-calibration}, parameterized calibration
is a solution to this issue, but is left for now as an area for future work.
Overall, this example further illustrates and confirms the ability
of the proposed method for inference, even in the case of multi-dimensional data where
direct approximations $p(\mathbf{x}|\theta_0)$ and $p(\mathbf{x}|\theta_1)$
are often difficult to construct.
